---
title: "Using available data function"
output: html_document
vignette: >
  %\VignetteIndexEntry{Using available data function}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

# Welcome to the Climate Future Toolbox's Available Data function

**This is an abridged version of the full available-data vignette. The vignette 
can be viewed in its entirety at https://github.com/earthlab/cft-CRAN/blob/master/full_vignettes/available-data.md**

This vignette provides a walk-through of a common use case of the Available Data function and the cft package:
understanding climate futures for a region of interest. 
We'll use Hot Springs National Park, located in Arkansas, USA and Yellowstone National Park, located in Wyoming, USA, as case studies.

Note that the available_data function is best used for downloading MACA climate model data for several climate variables from several climate models for any number of emission scenarios over a relatively small spatial region and over a relatively short time period.  If you would like to download MACA climate model data for several climate variables from several climate models and any number of emission scenarios in their entirety for a single lat/long location, you should use the single_point_firehose function in the cft pacakge.  A vignette on how to use the firehose function in the cft package is available at https://github.com/earthlab/cft/blob/main/vignettes/firehose.md.

### What you'll learn

This vignette will show you how to: 

- Access climate data for a spatial region of interest
- Produce a `data.frame` containing climate data
- Visualize historical and future data
- Generate and analyze new climate variables

### What you'll need

To get the most out of this vignette, we assume you have: 

- At least 500 MB of disk space
- Some familiarity with ggplot2
- Some familiarity with dplyr (e.g., [`filter()`](https://www.rdocumentation.org/packages/dplyr/versions/0.7.8/topics/filter), [`group_by()`](https://www.rdocumentation.org/packages/dplyr/versions/0.7.8/topics/group_by), and [`summarise()`](https://www.rdocumentation.org/packages/dplyr/versions/0.7.8/topics/summarise))

## About the data

Global Circulation Models (GCMs) provide estimates of historical and future 
climate conditions. 
The complexity of the climate system has lead to a large number GCMs and it is
common practice to examine outputs from many different models, treating each as 
one plausible future.

Most GCMs are spatially coarse (often 1 degree), but downscaling provides finer
scale estimates. The cft package uses one downscaled climate model called MACA 
(Multivariate Adaptive Climate Analog) Version 2 
([details here](https://www.climatologylab.org/maca.html)).

### Acquiring and subsetting data within National Park Service boundaries

This package was originally written with the National Park Service in mind, so
it has the option to use the name of any park (or monument, preserve, etc.) within
the NPS. Use the `cftdata()` function to specify a range of years, a set of models,
a set of parameters, and a set of representative concentration pathways to return. 
Leaving these arguments empty will result in a download of all available data 
for that location.

## Attach cft and check the list of available functions
```{r}
library(cft)
ls(pos="package:cft")
```

# Use read-only mode to find available data without initiating a full download.
```{r}
inputs <- cft::available_data()
```

## Look at the documentation for those functions
```{r}
?available_data
?single_point_firehose
```

Look at the variables, emission scenarios, and models for which data are available
```{r}
levels(as.factor(inputs$variable_names$Variable))
levels(as.factor(inputs$variable_names$`Variable abbreviation`))
levels(as.factor(inputs$variable_names$Scenario))
levels(as.factor(inputs$variable_names$`Scenario abbreviation`))
levels(as.factor(inputs$variable_names$Model))
levels(as.factor(inputs$variable_names$`Model abbreviation`))
```

This code downloads data for all models, all emission scenarios, and 5 climate variables.  Notice that this is a large download request.  
```{r}
library(magrittr)
library(dplyr)
input_variables <- inputs$variable_names %>% 
  filter(Variable %in% c("Maximum Relative Humidity", 
                       "Maximum Temperature", 
                       "Minimum Relative Humidity",          
                       "Minimum Temperature",                 
                       "Precipitation")) %>% 
  filter(Scenario %in% c( "RCP 4.5", "RCP 8.5")) %>% 
  filter(Model %in% c(
    "Beijing Climate Center - Climate System Model 1.1",
    "Beijing Normal University - Earth System Model",
    "Canadian Earth System Model 2",                                                                
  "Centre National de Recherches Météorologiques - Climate Model 5",                              
  "Commonwealth Scientific and Industrial Research Organisation - Mk3.6.0",                       
  "Community Climate System Model 4",                                                             
  "Geophysical Fluid Dynamics Laboratory - Earth System Model 2 Generalized Ocean Layer Dynamics",
  "Geophysical Fluid Dynamics Laboratory - Earth System Model 2 Modular Ocean",                   
  "Hadley Global Environment Model 2 - Climate Chemistry 365 (day) ",                             
 "Hadley Global Environment Model 2 - Earth System 365 (day)",                                   
 "Institut Pierre Simon Laplace (IPSL) - Climate Model 5A - Low Resolution",                     
 "Institut Pierre Simon Laplace (IPSL) - Climate Model 5A - Medium Resolution",                  
 "Institut Pierre Simon Laplace (IPSL) - Climate Model 5B - Low Resolution",                     
 "Institute of Numerical Mathematics Climate Model 4",                                           
 "Meteorological Research Institute - Coupled Global Climate Model 3",                           
 "Model for Interdisciplinary Research On Climate - Earth System Model",                         
 "Model for Interdisciplinary Research On Climate - Earth System Model - Chemistry",             
 "Model for Interdisciplinary Research On Climate 5",                                            
 "Norwegian Earth System Model 1 - Medium Resolution"  )) %>%
  
  pull("Available variable")
input_variables
```

# Establish area of interst (AOI) by bounding box
You will need to specify an appropriate area of interest for downloading spatial data. We utilize the functionality of open street map to make it fast and easy to download shapefiles for any area of interest. Here I am using the open street map (OSM) protocol to retrieve a shapefile for Hot Springs National Park. I choice this park because it is the smallest national park. 

This chunk of code is first using the [getbb()](https://www.rdocumentation.org/packages/osmdata/versions/0.1.9/topics/getbb) function to retrieve a bounding box matching a plain language search request. That bounding box is then fed into the API call that retrieves the data and converts it into an r-specific spatial data format called sf.

OSM uses a 'key' and 'value' system for organizing requests. You may want to spend some time familiarizing yourself with the immense library of data you can access with this system. https://wiki.openstreetmap.org/wiki/Tags
```{r bounding_box_small, cache=TRUE}
library(osmdata)
bb <- getbb("Hot Springs")
my_boundary <- opq(bb, timeout=300) %>% 
  add_osm_feature(key = "boundary", value = "national_park") %>% 
osmdata_sf() 
my_boundary
```

Here you re-calibrate the bounding box relative to the actual shapefiles you downloaded. The bounding box above was pulled from the osm database, this bounding box is for the polygons you actually used. These two boxes may be the same or they may be different but the one derived from your downloaded shapefile is reproducible. 

Notice that we specify osm_multipolygons instead of osm_polygons. This is a case-specific choice. When you download a shapefile from OSM, if will include a number of different spatial object types and you can choose several of those options to move forward. We chose multipolygons because they best matched our area of interest. Use the quick plot below to visualize your area of interest before continuing. 
```{r pulled_bounding_box_small, cache=TRUE}
library(osmdata)
library(sf)
boundaries <- my_boundary$osm_multipolygons
pulled_bb <- st_bbox(boundaries)
pulled_bb
```

This quick map allows you to check that the polygon you downloaded matches your expectations. 
```{r plot_of_area_of_interest_small, cache=TRUE, warning=FALSE, fig.height=8}
library(ggplot2)
basemap <- ggplot(data = boundaries) +
  geom_sf(fill = "cornflowerblue") +
  suppressWarnings(geom_sf_text(aes(label = boundaries$name))) 
basemap
```